A molecular dynamics study of the solvation of carbon dioxide in the ionic liquids \ce{[emim][B(CN)_4]} and \ce{[emim][NTf_2]}

### 2018-09-08 ionic_liquids.tex
Revision of the Computational Methods section

With the aim of setting up a computationally efficient protocol for the ILs, in this work we assess the possibility of modeling the ions as rigid bodies and employing the damped shifted force (DSF) method \cite{Fennell2006} for computing the electrostatic interactions.
As an alternative to the Ewald schemes \cite{Ewald_1921,Darden_1993,Hockney_1988}, McCann \textit{et al}. \cite{McCann_2013} tested the DSF method for various ILs modeled with the OPLS-AA force fields.
They concluded that using $\alpha$ = 0.2 {\AA}$^{-1}$ and a cutoff distance ($r_c$) of 15 {\AA} provide the best energetic agreement with the Ewald method.
In this work, comparisons are made in terms of the simulated densities and radial distribution functions for which, as will be shown in the next section, $\alpha$ = 0.2 {\AA}$^{-1}$ and $r_c$ = 12 {\AA} resulted in a good agreement.
These values were also employed for the charged solutes, with the exception of hexane, for which we applied undamped Coulombic interactions improving the prediction of the chemical potential.

The NPT algorithm of Kamberaj \textit{et al}. \cite{Kamberaj_2005} was used to perform the simulations with rigid bodies.
For flexible molecules, we applied the scheme due to Shinoda \textit{et al}. \cite{Shinoda2004} and C-H bonds were constrained by the SHAKE method \cite{Ryckaert1977}.
For both integrators, we used a time-step size ($\Delta t$) of 1 fs.

The gradual insertion/deletion of charged solutes involves scaling both the van der Waals (vdW) and the electrostatic intermolecular interactions.
In order to avoid numerical instabilities associated to the vdW scaling, a standard approach is to employ a soft-core potential with a nonlinear coupling.
In this paper we use the modified LJ potential due to Beutler \textit{et al}. \cite{Beutler_1994}.
The electrostatic interactions, on the other hand, can be handled by a linear scaling \cite{Naden_2015}.
Furthermore, the contributions to $\Delta G$ originating at the different couplings are taken into account separately, as follows:
\begin{equation}
\begin{split}
\Delta G_\text{sim} =& \; \Delta G_{\,\text{vdW}} (\lambda_{\,\text{vdW}} = 0\rightarrow  1, \lambda_{\, \text{Coul}} = 0) \\
+ \; &\Delta G_{\,\text{Coul}} (\lambda_{\, \text{vdW}} = 1,\lambda_{\,\text{Coul}} = 0 \rightarrow 1),
\end{split}
\end{equation}
where $\lambda_{\,\text{vdW}}$ and $\lambda_{\,\text{Coul}}$ are the coupling parameters of the soft-core and electrostatic potentials, respectively.
Note that in the process of coupling the vdW interactions, the Coulombic interactions are turned-off, while we set $\lambda_{\,\text{vdW}}$ = 1 when computing $\Delta G_{\,\text{Coul}}$.
This results in the ``end-point'' states given by ($\lambda_{\,\text{vdW}}$=0,$\lambda_{\,\text{Coul}}$=0) and ($\lambda_{\,\text{vdW}}$=1,$\lambda_{\,\text{Coul}}$=1), as expected.
The complete sets of the coupling parameters we use here are $\lambda_{\text{vdW}}$ = $\{0.0$, $0.05$, $0.1$, $0.2$, $0.25$, $0.3$, $0.35$, $0.4$, $0.45$, $0.5$, $0.6$, $0.7$, $0.8$, $0.9$, $1.0\}$ and $\lambda_{\text{Coul}}$ = $\{0.0$, $0.1$, $0.4$, $0.7$, $0.1\}$.

In this work, we carried out independent simulation at each state of $\lambda_{\text{Coul}}$, while the expanded ensemble (EE) method~\cite{Lyubartsev_1992} was applied for sampling the $\lambda_{\,\text{vdW}}$ set.
In short, an EE is constructed as a weighted sum of subensembles, each assuming a different value of the variable chosen to explore the free energy landscape, which in our case is $\lambda_{\,\text{vdW}}$.
This way, the sampling of an EE comprises both sampling each subensamble with fixed variables (N,P,T,$\lambda_{\,\text{vdW}}$) and transitions in $\lambda_{\,\text{vdW}}$.
Thus, a remarkable feature of the EE method is the possibility of exploring the whole alchemical route in a single simulation.
In this work, we perform a sequential sampling of the EE \cite{Chodera_2011_2}, which consists in sampling each subensamble via NPT molecular dynamics (MD) for a determined number of steps $N^{\text{EE}}$ followed by an attempt to change $\lambda_{\,\text{vdW}}$ using MC, and so on.
The $\lambda_{\,\text{vdW}}$ transitions are performed according to an independent sampling scheme \cite{Chodera_2011_2}.
In addition, the set of (arbitrary) weights $\eta$ of the subensambles enters in the acceptance criteria of MC, so it is possible to tune $\eta$ in order to enhance the sampling.
If differences $\eta_i$ - $\eta_j$ were closed to $\Delta G_{ji}$, we expect an uniform sampling in $\lambda_{\,\text{vdW}}$ \cite{Lyubartsev_1992}.
Of course, this can be achieved only iteratively.
The EE procedure, together with the \textit{soft-core} potential and the DSF method with linear coupling were implemented in the LAMMPS software package \cite{Plimpton1995}, which we used for carrying out all the simulations.

As demonstrated in Ref.~\citenum{Chodera_2011_2}, smaller values of $N^{EE}$ yield more decorrelated samples in each subensemble.
In this work, we employed $N^{EE}$ = 5 which, together with the specified $\lambda_{\, \text{vdW}}$ set, required up to 3 preliminary runs of the EE for determining the weights $\eta$ that yielded, in a final simulation, a ``sufficient'' sampling of each subensamble.
By ``sufficient'', we mean that the sampling provides an acceptable uncertainty in $\Delta G$.
The simulation times for the preliminary and final EE runs were 10 and 40 ns, respectively; and each $\lambda_{\text{Coul}}$ state was simulated for 10 ns.
Finally, the MBAR method \cite{Shirts_2008}, as implemented by Klimovich \textit{et al}. \cite{Klimovich_2015}, was applied to estimate $\Delta G_{\text{vdW}}$ and $\Delta G_{\text{Coul}}$.

The initial configurations were obtained with PLAYMOL \cite{playmol}, a software integrated with PACKMOL \cite{Mart_nez_2009}, using the tool \textit{build} to specify geometric data (bond and angle lengths, and dihedral angles), from which PLAYMOL automatically determines the Cartesian coordinates.
We first carried out NVT MD simulations of the isolated ion pairs in order to obtain a minimum energy conformation.
For this, we used the Nos\'{e}-Hoover thermostat \cite{Martyna1992} and a cubic box of length L = 50 {\AA} with non-periodic boundary conditions.
After a short run with $\Delta t$ = $0.5~\text{fs}$, we carried out a ``quenching'' simulation with $\Delta t$ = $1 ~\text{fs}$, in which the temperature is gradually reduced until reaching a value close to $0~\text{K}$.
The resulting coordinates were then specified in PLAYMOL \cite{playmol} to build the configurations of the liquid phases.

>>>>

With the aim of setting up a computationally efficient protocol for the IL simulations, in this work we assess the possibilities of modeling the ions as rigid bodies and employing the damped shifted force (DSF) method \cite{Fennell2006} for computing electrostatic interactions.
In this method, the electrostatic potential between two atoms with charges $q_i$ and $q_j$ interacting at a distance $r$ is given by
\begin{equation}
V_\text{\tiny DSF}(r) = V(r) - V(r_c) - V^\prime(r_c)(r-r_c),
\end{equation}
where $V(r) = q_i q_j {\text{erfc}(\alpha r)}/{r}$ is a damped Coulomb potential.
This potential is cut off at $r = r_c$, which makes it short-ranged and thus a cheaper alternative to Ewald-type schemes \cite{Ewald_1921,Darden_1993,Hockney_1988}.
McCann \textit{et al}. \cite{McCann_2013} tested the DSF method for various ILs modeled with the OPLS-AA force field.
They concluded that using $\alpha = 0.2~\text{\AA}^{-1}$ and $r_c = 15~\text{\AA}$ provides the best energetic agreement with the Ewald method.
In this work, detailed comparisons are made in terms of the simulated densities and radial distribution functions.
In Sec.~\ref{sec:results}, we show that by using the same $\alpha$ value but a considerably smaller cutoff distance ($r_c = 12 \text{\AA}$) one can obtain suitable results.
These values were employed here for the ILs as well as for solutes containing partially charged atoms.
The only exception was hexane, for which the use of undamped Coulombic interactions (i.e. $\alpha = 0$) resulted in improved chemical potential predictions.

The NPT algorithm of Kamberaj \textit{et al}. \cite{Kamberaj_2005} was used to integrate the equations of motion for rigid bodies.
For flexible molecules, we applied the integration scheme of Shinoda \textit{et al}. \cite{Shinoda2004}, with the lengths of all C-H bonds constrained by the SHAKE method \cite{Ryckaert1977}.
For both integrators, we used a small time-step size ($\Delta t$) of 1 fs in order to avoid discretization-induced artifacts \cite{Eastwood_2010, Silveira_2017}.

The gradual insertion/deletion of solutes involves scaling both the van der Waals (vdW) and the electrostatic intermolecular interactions.
To avoid numerical instabilities associated to the vdW scaling, it is usual to employ a soft-core potential with nonlinear coupling.
In this paper, we use the modified LJ potential proposed by Beutler \textit{et al}. \cite{Beutler_1994}, given by $V_\text{\tiny vdW}(r) = 4 \epsilon \lambda_\text{\tiny vdW} (s^{-2} - s^{-1})$, where $s = (r/\sigma)^6 + (1 - \lambda_\text{\tiny vdW})/2$.
The electrostatic interactions, on the other hand, can be simply multiplied by a coupling parameter $\lambda_\text{\tiny Coul}$ \cite{Naden_2015}.
Furthermore, the vdW and Coulomb contributions to the free energy difference are taken into account separately.
That is,
\begin{equation}
\Delta G_\text{sim} = \Delta G_\text{\tiny vdW} + \Delta G_\text{\tiny Coul},
\end{equation}
where $\Delta G_\text{\tiny vdW}$ is computed by varying $\lambda_\text{\tiny vdW} \in [0,1]$ while keeping $\lambda_\text{\tiny Coul} = 0$, whereas $\Delta G_\text{\tiny Coul}$ is computed by varying $\lambda_\text{\tiny Coul} \in [0,1]$ while keeping $\lambda_\text{\tiny vdW} = 1$.
The sets of coupling parameter values we use here in all alchemical simulations are $\lambda_{\text{\tiny vdW}}$ = $\{0.0$, $0.05$, $0.1$, $0.2$, $0.25$, $0.3$, $0.35$, $0.4$, $0.45$, $0.5$, $0.6$, $0.7$, $0.8$, $0.9$, $1.0\}$ and $\lambda_{\text{\tiny Coul}}$ = $\{0.0$, $0.1$, $0.4$, $0.7$, $0.1\}$.

In this work, we compute $\Delta G_\text{\tiny Coul}$ by carrying out an independent simulation of 10 ns at each state of $\lambda_{\text{\tiny Coul}}$.
In the end, we apply the Multistate Bennett Acceptance Ratio (MBAR) method \cite{Shirts_2008} by using the \textit{Alchemical Analysis} package \cite{Klimovich_2015}.

In the case of $\Delta G_\text{\tiny vdW}$, the Expanded Ensemble method \cite{Lyubartsev_1992} is applied for sampling the whole $\lambda_\text{\tiny vdW}$ set in a single simulation.
An expanded ensemble is constructed as a weighted sum of subensembles, each one assuming a different value of the parameter selected for exploring the free energy landscape, which in our case is $\lambda_\text{\tiny vdW}$.
In the method adopted here \cite{Chodera_2011_2}, simulating an expanded ensemble comprises doing molecular dynamics (MD) at fixed N, P, T, and $\lambda_\text{\tiny vdW}$ amid Monte Carlo (MC) transitions in the value of $\lambda_\text{\tiny vdW}$.
The number of MD steps between transitions is fixed and, at each transition, one of the $\lambda_\text{\tiny vdW}$ states is picked at random according to a corresponding set of probabilities \cite{Chodera_2011_2}.
In addition, a set of biasing weights is included in the MC acceptance criterion, thus making it possible to enhance the sampling procedure by iteratively tuning these weights until all $\lambda_\text{\tiny vdW}$ states are visited with equal frequency \cite{Lyubartsev_1992}.
Given that sample decorrelation is improved by having short MD runs between MC transitions \cite{Chodera_2011_2}, we used only 5 steps in each run.
In the iteration process for tuning the biasing weights, the simulation times for the preliminary and the final expanded ensemble rounds are 10 and 40 ns, respectively.
In this way, 3 iterations were sufficient in all cases for attaining, in the final round, an acceptable uncertainty in $\Delta G_\text{sim}$ estimated via MBAR \cite{Shirts_2008, Klimovich_2015}.

We first carried out MD simulations of isolated ion pairs in order to obtain minimum-energy conformations.
For this, we used a Nos\'{e}-Hoover chain thermostat \cite{Martyna1992} with set-point temperature gradually changing down to nearly $0~\text{K}$.
The resulting coordinates were then used with the PACKMOL algorithm \cite{Mart_nez_2009} via the Playmol package \cite{playmol} to obtain initial liquid-phase configurations.
The expanded ensemble procedure, the softcore potential, and the DSF method  with linear coupling were implemented by customizing the LAMMPS package \cite{Plimpton1995}, which we used for carrying out all the simulations in this work.
